In-silico molecular enrichment and clearance of the human intracranial space

### 2025-06-09 paper/main.tex
revision I changes by Marius  II

\subsection*{Perivascular flow shapes and accelerates molecular enrichment}
\label{sec:pvs_flow_results}

@@

We study the effect of such rapid PVS flow on intracranial molecular
enrichment by comparing the more conservative baseline model and a
high PVS flow model, where the former still includes the net flow
contributions from CSF production and the cardiac pulse wave, while
the latter additionally includes the ultraslow vasomotion
contribution. In both models, tracers were first observed at the basal
artery after 48 minutes (\Cref{fig:pvs_b}A), defined by their first-time arrival (FTA), the time at which the concentration first exceeds $0.1$ mmol/l. At all upstream locations along the middle cerebral arteries
(MCAs) and anterior cerebral artery (ACA), we found substantially
reduced FTAs with higher PVS flow (\Cref{fig:pvs_b}B), up to 2.5 hours
earlier in the left M2 segment of the MCA. PVS concentrations peaked
before the surrounding tissue (0:48h vs 1:12h at the MCA and 1:00h vs
1:48h at MCA2) with high PVS flow, while these peaks occurred nearly
simultaneously in the baseline model. The earlier appearance and the
delay in time-to-peak between the PVS and surrounding tissue clearly
indicates the directionality of tracer enrichment -- it first arrives
in the PVS and subsequently spreads into the tissue and SAS
(\Cref{fig:pvs_b}C, D), and especially so with higher PVS flow. These
observations also hold on an aggregated level: computing the total
amount of tracer in the PVS as a function of the distance to the
arterial network roots, we find accelerated tracer transport with
higher PVS flow, especially at earlier time points (2--9 hours)
(\Cref{fig:pvs_b}E). Moreover, differences in tracer transport along the
PVS translate to altered enrichment patterns on the whole-organ
scale. For instance, after 4-6 hours, the faster-moving tracer in the
PVS is clearly visible in the space adjacent to the PVS of the ACA, MCA, and other arteries (\Cref{fig:pvs_b}G). In contrast,
there are no clear signs of early enrichment surrounding the PVS in
the baseline model (\Cref{fig:pvs_b}F).

@@

\begin{figure}
    \centering
    \includegraphics[width=0.99\textwidth]{figures/figure3b.png}
    \caption{\textbf{Perivascular flow shapes and accelerates molecular enrichment.}
    A) Comparison of mean tracer concentration in the PVS and over the outer PVS surface at key locations of the arterial tree, for the baseline (lower) and high PVS flow (upper) models, with labels: ``peak'': time-to-peak (h), ``$\Delta t$'': time difference between peaks (h), and ``FTA'': first-time tracer arrival (h).
    B) first-time arrival for the main trunks of the periarterial network over the distance from the closest network root node (BA, ICA-R or ICA-L for the baseline, the high PVS flow and the high PVS dispersion models);}
    \label{fig:pvs_b}
\end{figure}
\begin{figure}
\ContinuedFloat
\caption{(\textbf{cont.})
    C, D) 2D slices showing zoom-in on the region surrounding the MCA-R and MCA2-R for the baseline (upper) and high PVS flow (lower) models. Innermost circle indicates the artery (blue), the surrounding annulus is the PVS (color represents concentration value), the surroundings are parenchyma and/or CSF spaces, separated by the pia (blue: towards the parenchyma, pink: towards the CSF));  
    E) Total amount of tracer in the periarterial spaces as a function of the distance to the nearest network root over time for the baseline (left) and the high PVS flow (right) models;
    F--H) sagittal (upper) and coronal (lower) views of the simulated tracer concentrations overlayed on the T1-weighted MR image after 4 and 6h for the baseline (K), high PVS flow (L) and high PVS dispersion (M) models (concentration opacity linearly increasing with concentration value, pial surface outlined in pink, arachnoid membrane in cyan, and arteries in dark red). An interactive visualization of the baseline and the high PVS flow model can be found \href{https://baseline-pvs-flow.surge.sh/}{here}, and \href{https://high-pvs-flow.surge.sh/}{here}.}
\end{figure}

@@

% Variations with sleep, exercise
Lifestyle factors such as sleep~\cite{xie2013sleep, miao2024brain,
  bojarskaite2023sleep, eide2021sleep, vinje2023human,
  hauglund2025norepinephrine, larsen2024sleep},
exercise~\cite{holstein2018voluntary, yildiz2022immediate}, and
alcohol intake~\cite{lundgaard2018beneficial} affect molecular
enrichment and clearance of the brain. In particular, pial arteries
display higher-amplitude low-frequency vasomotion during NREM and
intermediate sleep states, while REM sleep is associated with reduced
PVS width~\cite{bojarskaite2023sleep}. Peristaltic pumping is more
effective under both of these configurations, resulting in higher
estimated net PVS flow velocities~\cite{gjerde2023directional}. Even
higher PVS velocities would amplify the PVS flow effects simulated
here, with earlier arrival times along the PVSs, clear enrichment in
adjacent spaces, and accelerated molecular enrichment
overall. Moreover, sleep is linked with low-frequency oscillations in
human CSF flow~\cite{fultz2019coupled}. As we have shown here, the
level of dispersion in the CSF spaces strongly shapes molecular
enrichment patterns. While we, at baseline, account for the dispersion
induced by cardiac and respiratory pulsatility, the high dispersion
scenarios considered can be viewed as representative of additional
pulsatility induced by e.g.~neural waves~\cite{fultz2019coupled,
  williams2023neural} or other respiration patterns depending on
activity level. More broadly, we consider this as a key strength of
the current in-silico framework: allowing for integrating and
exploring the mechanistic macroscale implications of changes in
pulsatility at smaller spatial or temporal scales.

>>>>

\subsection*{Perivascular flow shapes and accelerates molecular transport}
\label{sec:pvs_flow_results}

@@

We study the effect of such rapid PVS flow on intracranial molecular
transport by comparing the more conservative baseline model and a
high PVS flow model, where the former still includes the net flow
contributions from CSF production and the cardiac pulse wave, while
the latter additionally includes the ultraslow vasomotion
contribution. In both models, tracers were first observed at the basal
artery after 48 minutes (\Cref{fig:pvs_b}A), defined by their first-time arrival (FTA), the time at which the concentration first exceeds $0.1$ mmol/l. At all upstream locations along the middle cerebral arteries
(MCAs) and anterior cerebral artery (ACA), we found substantially
reduced FTAs with higher PVS flow (\Cref{fig:pvs_b}B), up to 2.5 hours
earlier in the left M2 segment of the MCA. PVS concentrations peaked
before the surrounding tissue (0:48h vs 1:12h at the MCA and 1:00h vs
1:48h at MCA2) with high PVS flow, while these peaks occurred nearly
simultaneously in the baseline model. The earlier appearance and the
delay in time-to-peak between the PVS and surrounding tissue clearly
indicates the directionality of tracer enrichment -- it first arrives
in the PVS and subsequently spreads into the tissue and SAS
(\Cref{fig:pvs_b}C, D), and especially so with higher PVS flow. These
observations also hold on an aggregated level: computing the total
amount of tracer in the PVS as a function of the distance to the
arterial network roots, we find accelerated tracer transport with
higher PVS flow, especially at earlier time points (2--9 hours)
(\Cref{fig:pvs_b}E). Moreover, differences in tracer transport along the
PVS translate to altered enrichment patterns on the whole-organ
scale. For instance, after 4-6 hours, the faster-moving tracer in the
PVS is clearly visible in the space adjacent to the PVS of the ACA, MCA, and other arteries (\Cref{fig:pvs_b}G). In contrast,
there are no clear signs of early enrichment surrounding the PVS in
the baseline model (\Cref{fig:pvs_b}F).

@@

\begin{figure}
    \centering
    \includegraphics[width=0.99\textwidth]{figures/figure3b.png}
    \caption{\textbf{Perivascular flow shapes and accelerates molecular transport.}
    A) Comparison of mean tracer concentration in the PVS and over the outer PVS surface at key locations of the arterial tree, for the baseline (lower) and high PVS flow (upper) models, with labels: ``peak'': time-to-peak (h), ``$\Delta t$'': time difference between peaks (h), and ``FTA'': first-time tracer arrival (h).
    B) first-time arrival for the main trunks of the periarterial network over the distance from the closest network root node (BA, ICA-R or ICA-L for the baseline, the high PVS flow and the high PVS dispersion models);}
    \label{fig:pvs_b}
\end{figure}
\begin{figure}
\ContinuedFloat
\caption{(\textbf{cont.})
    C, D) 2D slices showing zoom-in on the region surrounding the MCA-R and MCA2-R for the baseline (upper) and high PVS flow (lower) models. Innermost circle indicates the artery (blue), the surrounding annulus is the PVS (color represents concentration value), the surroundings are parenchyma and/or CSF spaces, separated by the pia (blue: towards the parenchyma, pink: towards the CSF));  
    E) Total amount of tracer in the periarterial spaces as a function of the distance to the nearest network root over time for the baseline (left) and the high PVS flow (right) models;
    F--H) sagittal (upper) and coronal (lower) views of the simulated tracer concentrations overlayed on the T1-weighted MR image after 4 and 6h for the baseline (K), high PVS flow (L) and high PVS dispersion (M) models (concentration opacity linearly increasing with concentration value, pial surface outlined in pink, arachnoid membrane in cyan, and arteries in dark red). An interactive visualization of the baseline and the high PVS flow model can be found \href{https://baseline-pvs-flow.surge.sh/}{here}, and \href{https://high-pvs-flow.surge.sh/}{here}.}
\end{figure}

@@

% Variations with sleep, exercise
Lifestyle factors such as sleep~\cite{xie2013sleep, miao2024brain,
  bojarskaite2023sleep, eide2021sleep, vinje2023human,
  hauglund2025norepinephrine, larsen2024sleep},
exercise~\cite{holstein2018voluntary, yildiz2022immediate}, and
alcohol intake~\cite{lundgaard2018beneficial} affect molecular
enrichment and clearance of the brain. In particular, pial arteries
display higher-amplitude low-frequency vasomotion during NREM and
intermediate sleep states, while REM sleep is associated with reduced
PVS width~\cite{bojarskaite2023sleep}. Peristaltic pumping is more
effective under both of these configurations, resulting in higher
estimated net PVS flow velocities~\cite{gjerde2023directional}. Even
higher PVS velocities would amplify the PVS flow effects simulated
here, with earlier arrival times along the PVSs, clear enrichment in
adjacent spaces, and accelerated molecular enrichment
overall. Moreover, sleep is linked with low-frequency oscillations in
human CSF flow~\cite{fultz2019coupled}. As we have shown here, the
level of dispersion in the CSF spaces strongly shapes molecular
enrichment patterns. While we, at baseline, account for the dispersion
induced by cardiac and respiratory pulsatility, the high dispersion
scenarios considered can be viewed as representative of additional
pulsatility induced by e.g.~neural waves~\cite{fultz2019coupled,
  williams2023neural} or other respiration patterns depending on
activity level. More broadly, we consider this as a key strength of
the current in-silico framework: allowing for integrating and
exploring the mechanistic macroscale implications of changes in
pulsatility at smaller spatial or temporal scales.
This approach of modeling pulsatility through an effective dispersion factor is a simplification that does not capture the full complexity of transient fluid dynamics, for instance caused by non-uniform wall motion and ependymal cilia within the ventricular system~\cite{siyahhan2014flow}.
However, it is further supported by recent clinical observations. High-resolution MRI measurements by Hirschler et al.~\cite{hirschler2024region} revealed that CSF mobility in regions with high pulsatility is approximately tenfold greater than the self-diffusion of water, an enhancement factor consistent with the values computed in our model.
